---
title: "Exercise 1 - TMA4300 Computer Intensive Statistical Methods"
output: html_notebook
---

We first wnat to write an R function that generates  samples from a exponential distrobution with rate parameter $\lambda$. 
We now that the cumulative exponential distrobution takes the form $$ F(x;\lambda) =  1 - \lambda e ^{-\lambda x},  \quad x \ge 0 
$$

Computing the inverse cumulative function exploiting the uniform distrobution $U \sim Unif(0,1)$ can we get 
$$
X = F^{-1}(U) = - ln (U)/\lambda  
$$

We now want to show that this is true using simulation

```{r}
expdist <- function(rate, n) {
   x <-  -log( (1/rate)*( 1 - runif(n)))/rate
   return(x)
}

n <- 60000
lambda <- 1
plot(expdist(lambda,n))
plot(rexp(runif(n) ,rate=lambda ))

```




Let us consider a new example where the PDF has the form
$$
g(x) = 
\begin{cases}
cx^{\alpha -1},& \quad 0 < x < 1 \\
ce^{-x},& \quad x  \ge 1 \\
\end{cases}
$$

The cumulative function can be computed with integration

$$
G(x) = 
\begin{cases}
\frac{c}{\alpha}x^{\alpha},& \quad 0 < x < 1 \\
-c(e^{-x} - e^{-1}),& \quad x  \ge 1 \\
\end{cases}
$$
And similarly with the inverse cumulative function

$$
G^{-1} (y) = \begin{cases}
 \frac{1}{\alpha} ln(y \frac{\alpha}{c}), \quad 0<y < c/\alpha \\
-ln( e^{-1} - y/c) 
\end{cases}
$$
THis can similarly be generated such that 
```{r}
plot(cars)
```